---
title: '7\. Worksheet: Diversity Synthesis'
author: "Student Name; Z620: Quantitative Biodiversity, Indiana University"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=2.54cm
---
  
## OVERVIEW

In this worksheet, you will conduct exercises that reinforce fundamental concepts of biodiversity.
Specifically, you will construct a a site-by-species matrix by sampling confectionery taxa.
With this primary data structure, you will then answer questions and generate figures using tools from previous weeks, along with wrangling techniques that we learned about in class. 

## Directions:
1. In the Markdown version of this document in your cloned repo, change "Student Name" on line 3 (above) to your name.
2. Complete as much of the worksheet as possible during class.
3. Refer to previous handouts to help with developing of questions and writing of code.
4. Answer questions in the worksheet.
Space for your answer is provided in this document and indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio (color may vary if you changed the editor theme).
5. Before you leave the classroom, **push** this file to your GitHub repo.
6. For the assignment portion of the worksheet, follow the directions at the bottom of this file. 
7. When you are done, **Knit** the text and code into a PDF file.
8. After Knitting, submit the completed exercise by creating a **pull request** via GitHub.
Your pull request should include this file `7.DiversitySynthesis_Worskheet.Rmd` and the PDF output of `Knitr` (`DiversitySynthesis_Worskheet.pdf`).


## CONFECTIONARY EXERCISE GOALS

We will construct a site-by-species matrix using confectionery taxa (i.e, jelly beans). 
The instructors have created a **source community** with known abundance (*N*) and richness (*S*).
Like a real biological community, the species abundances are unevenly distributed such that a few jelly bean types are common while most are rare. 
Each student will sample the source community and separate their jelly beans into operational taxonomic units (OTUs).
With this information, we will create two data structures. 
First, we will create a site-by-species matrix. 
Second, we will make a taste-profile matrix. 
We will use these data structures to:

1) explore their sampling efforts and their effects on species richness using **coverage** and **rarefaction** concept,

2) measure **alpha diversity** for each sub-sample collated from data with their peers from the same source community,

3) examine **beta diversity** between each source community using the data generated across each source community, and 

4) use **data wrangling** tools they have learned during the class to accomplish the above goals.

## SAMPLING PROTOCOL TO CONSTRUCT A SITE-BY-SPECIES MATRIX

1. From the source community, each student should take a sample of ~75 individual or about half of a Dixie Cup.

2. At your desk, count the total number of individuals (*N*), sort the jelly beans into different types (i.e., OTUs), and quantify the abundance of each OTU.  

3. Work with other students in your group to assemble data into a site-by-species matrix on the white board. 
One person needs to create a worksheet (e.g., Google sheet) and share the group's site-by-species matrix with the class. 


## SAMPLING PROTOCOL TO CONSTRUCT A TASTE-PROFILE MATRIX

1. After creating the site-by-species matrix, each student will create a taste-profile based on the jelly beans contained in their own sample.
Pick your top 5-10 taxa based on flavor, color, sheen, speckle, etc. 

2. Work with other students in your group to assemble data into taste-profile incidence matrix on the white board where 1 = preferred and 0 = non-preferred jelly bean types.
One person needs to create a worksheet (e.g., Google sheet) and share the group's taste-profile matrix with the class. 
Make sure that each row (student) is identified by name and that each jelly bean type sampled is included with a unique taxonomic identifier. 


## DATA ANALYSIS

### 1) Sampling coverage and rarefaction curves

**Question 1:** Using the site-by-species matrix, explore how well your sampling effort was. 
Compare your sampling efforts with other groups. 
Do you think that your samples cover the actual diversity found in each source community? You can use rarefaction and coverage tools you have learned earlier. 

**Answer 1:** Use the space below to generate a rarefaction curve/sample coverage based on the data we collected in class for each community. 
Make sure to annotate your code using # symbols so others (including instructors) understand what you have done and why you have done it. 

```{r}
```


### 2) Alpha diversity

**Question 2:** Compare alpha diversity measures within sites and among communities. 
You can calculate and plot richness, Shannon diversity, and Simpson index. 
You can also convert these indices to effective number of species using the Hill numbers concept by generating a diversity profile, which will make comparisons easier across sites. 

What is the variation among the samples in your group and between the communities of other groups for the alpha diversity indices? Generate a hypothesis around the diversity metrics you chose and test your hypothesis. Interpret your findings.

**Answer 2a - Analysis:** Use the space below for code that is being used to analyze your data and test your hypotheses on your chosen alpha diversity tool. 
Make sure to annotate your code using # symbols so others (including instructors) understand what you have done and why you have done it.

```{r}
```

**Answer 2b - Plot:** With your analysis, create one (and only one, although it can have multiple panels) *publication-quality* figure.

```{r}
```

**Answer 2c - Interpret results:** Write an informative yet succinct (~5 sentences) caption that creates a "stand-alone" figure. Take a peek at figures and figure captions in a paper published in your favorite journal for inspiration.


### 3) Beta diversity

**Question 3:** Measure beta diversity using ordination and multivariate statistics methods. 
You can create a PCoA plot, based on Bray-Curtis dissimilarity, of sites and communities using different shape and color codes. 

```{r}
```

**Answer 3a - Plot:** With your analysis, create one (and only one, although it can have multiple panels) *publication-quality* figure.  

```{r}
```

**Answer 3b - Interpret results:** Write an informative yet succinct (~5 sentences) caption that creates a "stand-alone" figure. Take a peek at figures and figure captions in a paper published in your favorite journal for inspiration.


### 4) Taste profile

**Question 4** Using R, determine the most popular jelly bean in the class? 
**Question 5** Using R, quantitatively determine the other student in QB who has the most similar taste profile (i.e., "jelly buddy")
**Question 6** Create a cluster diagram or ordination depicting the taste profiles of the class with symbols labeled by student name. Make some observations. 




## SUBMITTING YOUR ASSIGNMENT
Use Knitr to create a PDF of your completed 7.DiversitySynthesis_Worksheet.Rmd document, push it to GitHub, and create a pull request.
Please make sure your updated repo includes both the pdf and RMarkdown files.

Unless otherwise noted, this assignment is due on **Wednesday, February 15^th^, 2023 at 12:00 PM (noon)**.